### 2024-06-18 nlp-papers/attention-is-all-you-need/ms.tex
Update paper study notes - Attention is All You Need

\documentclass{article}

% if you need to pass options to natbib, use, e.g.:
 \PassOptionsToPackage{numbers, compress}{natbib}
% before loading nips_2017
%
% to avoid loading the natbib package, add option nonatbib:
% \usepackage[nonatbib]{nips_2017}

% Keep final for now to remove annoying line numbers.
%\usepackage[final]{nips_2017}
% Keep final for now to remove annoying line numbers.

% to compile a camera-ready version, add the [final] option, e.g.:
\usepackage[final]{nips_2017}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage{fontspec}
\setmainfont{Ubuntu}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    anchorcolor=blue,
    urlcolor=blue,
    citecolor=red,
}
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{amsmath}
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{subfiles}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{enumerate}
\usepackage{subfiles}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{subfiles}

% % Custom Commands:
% \newcommand\blfootnote[1]{%
%   \begingroup
%   \renewcommand\thefootnote{}\footnote{#1}%
%   \addtocounter{footnote}{-1}%
%   \endgroup
% }

\newcommand\todo[1]{\textcolor{red}{[[#1]]}}
\newcommand\mc[1]{\mathcal{#1}}
\newcommand*\samethanks[1][\value{footnote}]{\footnotemark[#1]}
%keys for memory and values. Can be changed if needed
\newcommand{\kq}{q}
\newcommand{\km}{k}
\newcommand{\vq}{o}
\newcommand{\vm}{m}
\newcommand{\Wkq}{W_q}
\newcommand{\Wkm}{W_k}
\newcommand{\Wvq}{W_o}
\newcommand{\Wvm}{W_m}
\newcommand{\dmodel}{d_{\text{model}}}
\newcommand{\dffn}{d_{\text{ffn}}}
\newcommand{\dff}{d_{\text{ff}}}
\newcommand{\mbf}[1]{\mathbf{#1}}
%\newcommand{\kq}{{q}_k}
%\newcommand{\km}{{m}_k}
%\newcommand{\vq}{{q}_v}
%\newcommand{\vm}{{m}_v}
%\newcommand{\Wkq}{{W_q}_k}
%\newcommand{\Wkm}{{W_m}_k}
%\newcommand{\Wvq}{{W_q}_v}
%\newcommand{\Wvm}{{W_m}_v}
\newcommand\concat[3]{\left[#1 \parallel_#3 #2\right]}

\title{Attention Is All You Need}

% The \author macro works with any number of authors. There are two
% commands used to separate the names and addresses of multiple
% authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to
% break the lines. Using \AND forces a line break at that point. So,
% if LaTeX puts 3 of 4 authors names on the first line, and the last
% on the second line, try using \AND instead of \And before the third
% author name.
\author{
  \AND
  Ashish Vaswani\thanks{Equal contribution. Listing order is random. Jakob proposed replacing RNNs with self-attention and started the effort to evaluate this idea.
Ashish, with Illia, designed and implemented the first Transformer models and has been crucially involved in every aspect of this work. Noam proposed scaled dot-product attention, multi-head attention and the parameter-free position representation and became the other person involved in nearly every detail. Niki designed, implemented, tuned and evaluated countless model variants in our original codebase and tensor2tensor. Llion also experimented with novel model variants, was responsible for our initial codebase, and efficient inference and visualizations. Lukasz and Aidan spent countless long days designing various parts of and implementing tensor2tensor, replacing our earlier codebase, greatly improving results and massively accelerating our research.
}\\
  Google Brain\\
  \texttt{avaswani@google.com}\\
  \And
  Noam Shazeer\footnotemark[1]\\
  Google Brain\\
  \texttt{noam@google.com}\\
  \And
  Niki Parmar\footnotemark[1]\\
  Google Research\\
  \texttt{nikip@google.com}\\  
  \And
  Jakob Uszkoreit\footnotemark[1]\\
  Google Research\\
  \texttt{usz@google.com}\\
  \And  
  Llion Jones\footnotemark[1]\\
  Google Research\\
  \texttt{llion@google.com}\\   
  \And
  Aidan N. Gomez\footnotemark[1] \hspace{1.7mm}\thanks{Work performed while at Google Brain.}\\
  University of Toronto\\
  \texttt{aidan@cs.toronto.edu}
  \And
  {\L}ukasz Kaiser\footnotemark[1]\\
  Google Brain\\
  \texttt{lukaszkaiser@google.com}\\
  \And
  Illia Polosukhin\footnotemark[1]\hspace{1.7mm} \thanks{Work performed while at Google Research.}\\
  \texttt{illia.polosukhin@gmail.com}\\  
}

\begin{document}
\begin{center}
    \color{red}
    \large Provided proper attribution is provided, Google hereby grants permission to reproduce the tables and figures in this paper solely for use in journalistic or scholarly works.
\end{center}

\maketitle

\begin{abstract}
The dominant sequence transduction models are based on complex recurrent or convolutional neural networks that include an encoder and a decoder. The best performing models also connect the encoder and decoder through an attention mechanism. We propose a new simple network architecture, the Transformer, based solely on attention mechanisms, dispensing with recurrence and convolutions entirely. Experiments on two machine translation tasks show these models to be superior in quality while being  more parallelizable and requiring significantly less time to train. Our model achieves 28.4 BLEU on the WMT 2014 English-to-German translation task, improving over the existing best results, including ensembles, by over 2 BLEU.  On the WMT 2014 English-to-French translation task, our model establishes a new single-model state-of-the-art BLEU score of 41.8 after training for 3.5 days on eight GPUs, a small fraction of the training costs of the best models from the literature. We show that the Transformer generalizes well to other tasks by applying it successfully to English constituency parsing  both with large and limited training data.
% \blfootnote{Code available at \url{https://github.com/tensorflow/tensor2tensor}}

%TODO(noam): update results for new models.

%llion@: FAIR's paper seems to concentrate solely on the convolutional aspect of their model and have the attention as an after thought almost, this gives us a good opportunity to differentiate ourselves from their paper.

%We are simpler in a number of ways and should have the simplicity as a big selling point:
%\begin{itemize}
%\item No convolutions
%\item No need for such careful initializations and %normalization.
%\item Simpler non-lineararities, they use the gated linear %units.
%\item Less layers?
%\end{itemize}
%One thing we do more is that we have self attention.
%Another selling point is the increased interpretability as %shown with the visualizations. Which comes from the %simplicity and use of only attentions.
\end{abstract}

\section{Introduction}

\input{introduction}

\section{Background}

\input{background}

\section{Model Architecture}
\input{model_architecture}

\section{Why Self-Attention}
\input{why_self_attention}

\section{Training}
\input{training}

\section{Results} \label{sec:results}
\input{results}

\section{Conclusion}
In this work, we presented the Transformer, the first sequence transduction model based entirely on attention, replacing the recurrent layers most commonly used in encoder-decoder architectures with multi-headed self-attention.

For translation tasks, the Transformer can be trained significantly faster than architectures based on recurrent or convolutional layers. On both WMT 2014 English-to-German and WMT 2014 English-to-French translation tasks, we achieve a new state of the art. In the former task our best model outperforms even all previously reported ensembles. %We also provide an indication of the broader applicability of our models through experiments on English constituency parsing.

We are excited about the future of attention-based models and plan to apply them to other tasks. We plan to extend the Transformer to problems involving input and output modalities other than text and to investigate local, restricted attention mechanisms to efficiently handle large inputs and outputs such as images, audio and video.
Making generation less sequential is another research goals of ours.

The code we used to train and evaluate our models is available at \url{https://github.com/tensorflow/tensor2tensor}.

\paragraph{Acknowledgements} We are grateful to Nal Kalchbrenner and Stephan Gouws for
their fruitful comments, corrections and inspiration.

\bibliographystyle{plain}
%\bibliography{deeplearn}
\begin{thebibliography}{10}

\bibitem{layernorm2016}
Jimmy~Lei Ba, Jamie~Ryan Kiros, and Geoffrey~E Hinton.
\newblock Layer normalization.
\newblock {\em arXiv preprint arXiv:1607.06450}, 2016.

\bibitem{bahdanau2014neural}
Dzmitry Bahdanau, Kyunghyun Cho, and Yoshua Bengio.
\newblock Neural machine translation by jointly learning to align and
  translate.
\newblock {\em CoRR}, abs/1409.0473, 2014.

\bibitem{DBLP:journals/corr/BritzGLL17}
Denny Britz, Anna Goldie, Minh{-}Thang Luong, and Quoc~V. Le.
\newblock Massive exploration of neural machine translation architectures.
\newblock {\em CoRR}, abs/1703.03906, 2017.

\bibitem{cheng2016long}
Jianpeng Cheng, Li~Dong, and Mirella Lapata.
\newblock Long short-term memory-networks for machine reading.
\newblock {\em arXiv preprint arXiv:1601.06733}, 2016.

\bibitem{cho2014learning}
Kyunghyun Cho, Bart van Merrienboer, Caglar Gulcehre, Fethi Bougares, Holger
  Schwenk, and Yoshua Bengio.
\newblock Learning phrase representations using rnn encoder-decoder for
  statistical machine translation.
\newblock {\em CoRR}, abs/1406.1078, 2014.

\bibitem{xception2016}
Francois Chollet.
\newblock Xception: Deep learning with depthwise separable convolutions.
\newblock {\em arXiv preprint arXiv:1610.02357}, 2016.

\bibitem{gruEval14}
Junyoung Chung, {\c{C}}aglar G{\"{u}}l{\c{c}}ehre, Kyunghyun Cho, and Yoshua
  Bengio.
\newblock Empirical evaluation of gated recurrent neural networks on sequence
  modeling.
\newblock {\em CoRR}, abs/1412.3555, 2014.

\bibitem{dyer-rnng:16}
Chris Dyer, Adhiguna Kuncoro, Miguel Ballesteros, and Noah~A. Smith.
\newblock Recurrent neural network grammars.
\newblock In {\em Proc. of NAACL}, 2016.

\bibitem{JonasFaceNet2017}
Jonas Gehring, Michael Auli, David Grangier, Denis Yarats, and Yann~N. Dauphin.
\newblock Convolutional sequence to sequence learning.
\newblock {\em arXiv preprint arXiv:1705.03122v2}, 2017.

\bibitem{graves2013generating}
Alex Graves.
\newblock Generating sequences with recurrent neural networks.
\newblock {\em arXiv preprint arXiv:1308.0850}, 2013.

\bibitem{he2016deep}
Kaiming He, Xiangyu Zhang, Shaoqing Ren, and Jian Sun.
\newblock Deep residual learning for image recognition.
\newblock In {\em Proceedings of the IEEE Conference on Computer Vision and
  Pattern Recognition}, pages 770--778, 2016.

\bibitem{hochreiter2001gradient}
Sepp Hochreiter, Yoshua Bengio, Paolo Frasconi, and J{\"u}rgen Schmidhuber.
\newblock Gradient flow in recurrent nets: the difficulty of learning long-term
  dependencies, 2001.

\bibitem{hochreiter1997}
Sepp Hochreiter and J{\"u}rgen Schmidhuber.
\newblock Long short-term memory.
\newblock {\em Neural computation}, 9(8):1735--1780, 1997.

\bibitem{huang-harper:2009:EMNLP}
Zhongqiang Huang and Mary Harper.
\newblock Self-training {PCFG} grammars with latent annotations across
  languages.
\newblock In {\em Proceedings of the 2009 Conference on Empirical Methods in
  Natural Language Processing}, pages 832--841. ACL, August 2009.

\bibitem{jozefowicz2016exploring}
Rafal Jozefowicz, Oriol Vinyals, Mike Schuster, Noam Shazeer, and Yonghui Wu.
\newblock Exploring the limits of language modeling.
\newblock {\em arXiv preprint arXiv:1602.02410}, 2016.

\bibitem{extendedngpu}
{\L}ukasz Kaiser and Samy Bengio.
\newblock Can active memory replace attention?
\newblock In {\em Advances in Neural Information Processing Systems, ({NIPS})},
  2016.

\bibitem{neural_gpu}
\L{}ukasz Kaiser and Ilya Sutskever.
\newblock Neural {GPU}s learn algorithms.
\newblock In {\em International Conference on Learning Representations
  ({ICLR})}, 2016.

\bibitem{NalBytenet2017}
Nal Kalchbrenner, Lasse Espeholt, Karen Simonyan, Aaron van~den Oord, Alex
  Graves, and Koray Kavukcuoglu.
\newblock Neural machine translation in linear time.
\newblock {\em arXiv preprint arXiv:1610.10099v2}, 2017.

\bibitem{structuredAttentionNetworks}
Yoon Kim, Carl Denton, Luong Hoang, and Alexander~M. Rush.
\newblock Structured attention networks.
\newblock In {\em International Conference on Learning Representations}, 2017.

\bibitem{kingma2014adam}
Diederik Kingma and Jimmy Ba.
\newblock Adam: A method for stochastic optimization.
\newblock In {\em ICLR}, 2015.

\bibitem{Kuchaiev2017Factorization}
Oleksii Kuchaiev and Boris Ginsburg.
\newblock Factorization tricks for {LSTM} networks.
\newblock {\em arXiv preprint arXiv:1703.10722}, 2017.

\bibitem{lin2017structured}
Zhouhan Lin, Minwei Feng, Cicero Nogueira~dos Santos, Mo~Yu, Bing Xiang, Bowen
  Zhou, and Yoshua Bengio.
\newblock A structured self-attentive sentence embedding.
\newblock {\em arXiv preprint arXiv:1703.03130}, 2017.

\bibitem{multiseq2seq}
Minh-Thang Luong, Quoc~V. Le, Ilya Sutskever, Oriol Vinyals, and Lukasz Kaiser.
\newblock Multi-task sequence to sequence learning.
\newblock {\em arXiv preprint arXiv:1511.06114}, 2015.

\bibitem{luong2015effective}
Minh-Thang Luong, Hieu Pham, and Christopher~D Manning.
\newblock Effective approaches to attention-based neural machine translation.
\newblock {\em arXiv preprint arXiv:1508.04025}, 2015.

\bibitem{marcus1993building}
Mitchell~P Marcus, Mary~Ann Marcinkiewicz, and Beatrice Santorini.
\newblock Building a large annotated corpus of english: The penn treebank.
\newblock {\em Computational linguistics}, 19(2):313--330, 1993.

\bibitem{mcclosky-etAl:2006:NAACL}
David McClosky, Eugene Charniak, and Mark Johnson.
\newblock Effective self-training for parsing.
\newblock In {\em Proceedings of the Human Language Technology Conference of
  the NAACL, Main Conference}, pages 152--159. ACL, June 2006.

\bibitem{decomposableAttnModel}
Ankur Parikh, Oscar Täckström, Dipanjan Das, and Jakob Uszkoreit.
\newblock A decomposable attention model.
\newblock In {\em Empirical Methods in Natural Language Processing}, 2016.

\bibitem{paulus2017deep}
Romain Paulus, Caiming Xiong, and Richard Socher.
\newblock A deep reinforced model for abstractive summarization.
\newblock {\em arXiv preprint arXiv:1705.04304}, 2017.

\bibitem{petrov-EtAl:2006:ACL}
Slav Petrov, Leon Barrett, Romain Thibaux, and Dan Klein.
\newblock Learning accurate, compact, and interpretable tree annotation.
\newblock In {\em Proceedings of the 21st International Conference on
  Computational Linguistics and 44th Annual Meeting of the ACL}, pages
  433--440. ACL, July 2006.

\bibitem{press2016using}
Ofir Press and Lior Wolf.
\newblock Using the output embedding to improve language models.
\newblock {\em arXiv preprint arXiv:1608.05859}, 2016.

\bibitem{sennrich2015neural}
Rico Sennrich, Barry Haddow, and Alexandra Birch.
\newblock Neural machine translation of rare words with subword units.
\newblock {\em arXiv preprint arXiv:1508.07909}, 2015.

\bibitem{shazeer2017outrageously}
Noam Shazeer, Azalia Mirhoseini, Krzysztof Maziarz, Andy Davis, Quoc Le,
  Geoffrey Hinton, and Jeff Dean.
\newblock Outrageously large neural networks: The sparsely-gated
  mixture-of-experts layer.
\newblock {\em arXiv preprint arXiv:1701.06538}, 2017.

\bibitem{srivastava2014dropout}
Nitish Srivastava, Geoffrey~E Hinton, Alex Krizhevsky, Ilya Sutskever, and
  Ruslan Salakhutdinov.
\newblock Dropout: a simple way to prevent neural networks from overfitting.
\newblock {\em Journal of Machine Learning Research}, 15(1):1929--1958, 2014.

\bibitem{sukhbaatar2015}
Sainbayar Sukhbaatar, Arthur Szlam, Jason Weston, and Rob Fergus.
\newblock End-to-end memory networks.
\newblock In C.~Cortes, N.~D. Lawrence, D.~D. Lee, M.~Sugiyama, and R.~Garnett,
  editors, {\em Advances in Neural Information Processing Systems 28}, pages
  2440--2448. Curran Associates, Inc., 2015.

\bibitem{sutskever14}
Ilya Sutskever, Oriol Vinyals, and Quoc~VV Le.
\newblock Sequence to sequence learning with neural networks.
\newblock In {\em Advances in Neural Information Processing Systems}, pages
  3104--3112, 2014.

\bibitem{DBLP:journals/corr/SzegedyVISW15}
Christian Szegedy, Vincent Vanhoucke, Sergey Ioffe, Jonathon Shlens, and
  Zbigniew Wojna.
\newblock Rethinking the inception architecture for computer vision.
\newblock {\em CoRR}, abs/1512.00567, 2015.

\bibitem{KVparse15}
{Vinyals {\&} Kaiser}, Koo, Petrov, Sutskever, and Hinton.
\newblock Grammar as a foreign language.
\newblock In {\em Advances in Neural Information Processing Systems}, 2015.

\bibitem{wu2016google}
Yonghui Wu, Mike Schuster, Zhifeng Chen, Quoc~V Le, Mohammad Norouzi, Wolfgang
  Macherey, Maxim Krikun, Yuan Cao, Qin Gao, Klaus Macherey, et~al.
\newblock Google's neural machine translation system: Bridging the gap between
  human and machine translation.
\newblock {\em arXiv preprint arXiv:1609.08144}, 2016.

\bibitem{DBLP:journals/corr/ZhouCWLX16}
Jie Zhou, Ying Cao, Xuguang Wang, Peng Li, and Wei Xu.
\newblock Deep recurrent models with fast-forward connections for neural
  machine translation.
\newblock {\em CoRR}, abs/1606.04199, 2016.

\bibitem{zhu-EtAl:2013:ACL}
Muhua Zhu, Yue Zhang, Wenliang Chen, Min Zhang, and Jingbo Zhu.
\newblock Fast and accurate shift-reduce constituent parsing.
\newblock In {\em Proceedings of the 51st Annual Meeting of the ACL (Volume 1:
  Long Papers)}, pages 434--443. ACL, August 2013.

\end{thebibliography}
%\newpage
\input{visualizations}
%\appendix
%\newpage
%\input{parameter_attention}

%\input{sqrt_d_trick}

\end{document}

### 2024-07-05 nlp-papers/reasoning-via-planning/emnlp2023.tex
Update paper study notes - Attention is All You Need

\documentclass[11pt]{article}
\usepackage{EMNLP2023}
\usepackage{times}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{listings}
\lstset{
  backgroundcolor=\color{gray!8},
  basicstyle=\ttfamily,
  columns=fullflexible
}
% For proper rendering and hyphenation of words containing Latin characters (including in bib files)
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{inconsolata}
\usepackage{xspace}
\usepackage{booktabs}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsmath, amsfonts, amssymb}

\usepackage{makecell}
\usepackage[normalem]{ulem}
\usepackage{pifont}% http://ctan.org/pkg/pifont
\newcommand{\cmark}{{\ding{51}}}
\newcommand{\xmark}{{\ding{55}}}
\NewDocumentCommand{\hsb}{ mO{} }{\textcolor{pink}{\textsuperscript{\textit{Shibo}}\textsf{\textbf{\small[#1]}}}}

\NewDocumentCommand{\gy}{ mO{} }{\textcolor{purple}{\textsuperscript{\textit{Yi}}\textsf{\textbf{\small[#1]}}}}

\NewDocumentCommand{\hzt}{ mO{} }{\textcolor{red}{\textsuperscript{\textit{ZT}}\textsf{\textbf{\small[#1]}}}}

\NewDocumentCommand{\zhen}{ mO{} }{\textcolor{blue}{\textsuperscript{\textit{ZW}}\text{\text{\small[#1]}}}}

\NewDocumentCommand{\jjh}{ mO{} }{\textcolor{cyan}{\textsuperscript{\textit{Joshua}}\text{\text{\small[#1]}}}}

\newcommand{\dummyfig}[1]{
  \centering
  \fbox{
    \begin{minipage}[c][0.3\textheight][c]{0.5\textwidth}
      \centering{#1}
    \end{minipage}
  }
}

\def\blocksworld{{Blocksworld}\xspace}
\def\ours{{RAP}\xspace}

% If the title and author information does not fit in the area allocated, uncomment the following
%
%\setlength\titlebox{<dim>}
%
% and set <dim> to something 5cm or larger.

\title{Reasoning with Language Model is
Planning with World Model}

% Author information can be set in various styles:
% For several authors from the same institution:
% \author{Author 1 \and ... \and Author n \\
%         Address line \\ ... \\ Address line}
% if the names do not fit well on one line use
%         Author 1 \\ {\bf Author 2} \\ ... \\ {\bf Author n} \\
% For authors from different institutions:
% \author{Author 1 \\ Address line \\  ... \\ Address line
%         \And  ... \And
%         Author n \\ Address line \\ ... \\ Address line}
% To start a separate ``row'' of authors use \AND, as in
% \author{Author 1 \\ Address line \\  ... \\ Address line
%         \AND
%         Author 2 \\ Address line \\ ... \\ Address line \And
%         Author 3 \\ Address line \\ ... \\ Address line}

\author{%
Shibo Hao\textsuperscript{$*\clubsuit$} \ Yi Gu\thanks{\, equal contribution}\textsuperscript{$*\clubsuit$} \
Haodi Ma\textsuperscript{$\diamondsuit$}\  Joshua Jiahua Hong\textsuperscript{$\clubsuit$} \\ \textbf{Zhen Wang\textsuperscript{$\clubsuit$ $\spadesuit$} \
Daisy Zhe Wang\textsuperscript{$\diamondsuit$}\ Zhiting Hu\textsuperscript{$\clubsuit$}}  \vspace{5pt} \\
\textsuperscript{$\clubsuit$}UC San Diego, \textsuperscript{$\diamondsuit$}University of Florida\\
\textsuperscript{$\spadesuit$}Mohamed bin Zayed University of Artificial Intelligence \\
\texttt{\{s5hao, yig025, jjhong, zhw085, zhh019\}@ucsd.edu}  \\
\texttt{\{ma.haodi, daisyw\}@ufl.edu}}

% \author{First Author \\
%   Affiliation / Address line 1 \\
%   Affiliation / Address line 2 \\
%   Affiliation / Address line 3 \\
%   \texttt{email@domain} \\\And
%   Second Author \\
%   Affiliation / Address line 1 \\
%   Affiliation / Address line 2 \\
%   Affiliation / Address line 3 \\
%   \texttt{email@domain} \\}

\begin{document}
\maketitle
\begin{abstract}
Large language models (LLMs) have shown remarkable reasoning capabilities, particularly with chain-of-thought (CoT) prompting. However, LLMs sometimes still struggle with problems that are easy for humans, such as generating action plans to achieve given goals in an environment, or performing complex math or logical reasoning. The deficiency stems from the key fact that LLMs lack an internal \emph{world model} to predict the world \emph{state} (e.g., environment status, intermediate variable values) and simulate long-term outcomes of actions. This prevents LLMs from performing deliberate planning akin to human brains, which involves exploring alternative reasoning paths, anticipating future states and rewards, and iteratively refining existing reasoning steps. To overcome the limitations, we propose a new LLM reasoning framework, {\bf \underline{R}easoning vi\underline{a} \underline{P}lanning (RAP)}. RAP repurposes the LLM as both a world model and a reasoning agent, and incorporates a principled planning algorithm based on Monte Carlo Tree Search for strategic exploration in the vast reasoning space. During reasoning, the LLM (as agent) incrementally builds a reasoning tree under the guidance of the LLM (as world model) and rewards, and efficiently obtains a high-reward reasoning path with a proper balance between exploration \emph{vs.} exploitation.
% As a general reasoning framework,
We apply RAP to various challenging reasoning problems including plan generation, math reasoning, and logical inference, and demonstrate its superiority over strong baselines. RAP with LLaMA-33B even surpasses CoT with GPT-4, achieving 33\% relative improvement in a plan generation setting.\footnote{The code is available at \url{https://github.com/Ber666/llm-reasoners}}
\end{abstract}
\input{sections/intro}
\input{sections/related_works}
\input{sections/method}
\input{sections/experiment}
\input{sections/conclusion}
\section*{Limitations}
In this work, we mainly focus on utilizing frozen LLMs, whose abilities might be bounded by the pre-training. In the future, it is worth exploring how to fine-tune LLMs to better reason and serve as a world model \cite{xiang2023language}, as well as how to combine external tools \cite{hao2023toolkengpt,schick2023toolformer} with RAP to solve more complex real-world problems.

\section*{Ethics Statement}
In this paper, we primarily focus on the applications on
plan generation, mathematical reasoning, and logical reasoning, posing no significant ethical concerns. We recognize that future research on border applications of reasoning with LLMs may pose a risk of misuse, and we recommend careful consideration of all aspects of safety before relevant techniques are applied to the real world.
%Scientific work published at EMNLP 2023 must comply with the \href{https://www.aclweb.org/portal/content/acl-code-ethics}{ACL Ethics Policy}. We encourage all authors to include an explicit ethics statement on the broader impact of the work, or other ethical considerations after the conclusion but before the references. The ethics statement will not count toward the page limit (8 pages for long, 4 pages for short papers).

% Entries for the entire Anthology, followed by custom entries
\bibliography{main}
\bibliographystyle{acl_natbib}

\appendix
\input{sections/appendix}

\end{document}

### 2024-07-05 nlp-papers/reasoning-via-planning/emnlp2023.tex
deploy: 3dc9494d6b5c6fe6e828db8f0e990c1af8bb5bcf
deploy: 09732c01d7193d9bf0f6a417c56d46d1862db069

@@
 \usepackage[T1]{fontenc}
+\usepackage{fontspec}
+\setmainfont{Ubuntu}
+\setmonofont{Ubuntu}
+\setsansfont{Ubuntu}
+\urlstyle{same}
+\usepackage[most]{tcolorbox}
+\newtcbox{\roundinlinebox}[1][red]{
+    on line,
+    arc=7pt,
+    colback=#1!10!white,
+    colframe=#1!50!white,
+    before upper={\rule[-3pt]{0pt}{10pt}},
+    boxrule=1pt,
+    boxsep=0pt,left=6pt,right=6pt,top=2pt,bottom=2pt
+}
 \usepackage[utf8]{inputenc}
@@
 \def\ours{{RAP}\xspace}
+
+\usepackage{soul}
+\usepackage{tikz}
+\usetikzlibrary{calc}
+\usetikzlibrary{decorations.pathmorphing}
+
+\makeatletter
+
+\newcommand{\defhighlighter}[3][]{%
+    \tikzset{every highlighter/.style={color=#2, fill opacity=#3, #1}}%
+}
+
+\defhighlighter{yellow}{.5}
+
+\newcommand{\highlight@DoHighlight}{
+    \fill [ decoration = {random steps, amplitude=1pt, segment length=15pt}
+        , outer sep = -15pt, inner sep = 0pt, decorate
+        , every highlighter, this highlighter ]
+    ($(begin highlight)+(0,8pt)$) rectangle ($(end highlight)+(0,-3pt)$) ;
+}
+
+\newcommand{\highlight@BeginHighlight}{
+    \coordinate (begin highlight) at (0,0) ;
+}
+
+\newcommand{\highlight@EndHighlight}{
+    \coordinate (end highlight) at (0,0) ;
+}
+
+\newdimen\highlight@previous
+\newdimen\highlight@current
+
+\DeclareRobustCommand*\highlight[1][]{%
+    \tikzset{this highlighter/.style={#1}}%
+    \SOUL@setup
+%
+    \def\SOUL@preamble{%
+        \begin{tikzpicture}[overlay, remember picture]
+            \highlight@BeginHighlight
+            \highlight@EndHighlight
+        \end{tikzpicture}%
+    }%
+%
+    \def\SOUL@postamble{%
+        \begin{tikzpicture}[overlay, remember picture]
+            \highlight@EndHighlight
+            \highlight@DoHighlight
+        \end{tikzpicture}%
+    }%
+%
+    \def\SOUL@everyhyphen{%
+        \discretionary{%
+            \SOUL@setkern\SOUL@hyphkern
+            \SOUL@sethyphenchar
+            \tikz[overlay, remember picture] \highlight@EndHighlight ;%
+        }{%
+        }{%
+            \SOUL@setkern\SOUL@charkern
+        }%
+    }%
+%
+    \def\SOUL@everyexhyphen##1{%
+        \SOUL@setkern\SOUL@hyphkern
+        \hbox{##1}%
+        \discretionary{%
+            \tikz[overlay, remember picture] \highlight@EndHighlight ;%
+        }{%
+        }{%
+            \SOUL@setkern\SOUL@charkern
+        }%
+    }%
+%
+    \def\SOUL@everysyllable{%
+        \begin{tikzpicture}[overlay, remember picture]
+            \path let \p0 = (begin highlight), \p1 = (0,0) in \pgfextra
+            \global\highlight@previous=\y0
+            \global\highlight@current =\y1
+            \endpgfextra (0,0) ;
+            \ifdim\highlight@current < \highlight@previous
+            \highlight@DoHighlight
+            \highlight@BeginHighlight
+            \fi
+        \end{tikzpicture}%
+        \the\SOUL@syllable
+        \tikz[overlay, remember picture] \highlight@EndHighlight ;%
+    }%
+    \SOUL@
+}
+\makeatother
 

### 2024-07-05 nlp-papers/reasoning-via-planning/sections/method.tex
deploy: 09732c01d7193d9bf0f6a417c56d46d1862db069

% \vspace{-5pt}
\section{Reasoning via Planning (RAP)}
% \vspace{-5pt}

% Most recent LLMs use an autoregressive decoder architecture when solving reasoning tasks, which prevents them from looking ahead and doing exploration when making a reasoning step.
% This inability may be blamed for their failures on complex multi-step reasoning tasks.
% To address these shortages, we propose to form reasoning as a planning task. In this section, we will introduce the world model formulation (how to look ahead?), and the search algorithm (how to do exploration?). \gy{modify this paragraph ; we do not want to use look ahead}

% During the process of solving complex reasoning problems, humans frequently engage in planning by simulating each step of the reasoning process using the world model in their minds.
% They iterate through recognizing problematic steps and exploring alternative reasoning steps to enhance the overall reasoning path.
% % \gy{unsure whether the description is accurate}
% However, existing approaches that employ LLMs to tackle reasoning problems solely rely on autoregressive generation of the reasoning trace, preventing them from recovering from incorrect early steps and exploring alternative paths.

% In order to overcome the limitations arising from this discrepancy,

% \subsection{Planning with Monte Carlo Tree Search} \label{sec:mcts}

In this section, we present the Reasoning via Planning (RAP) framework that enables LLMs to strategically plan a coherent reasoning trace for solving a wide range of reasoning tasks. We first \roundinlinebox[green]{build the world model} by repurposing the LLM with prompting (Section~\ref{sec:formulation}). The world model serves as the foundation for deliberate planning, by allowing the LLM to plan ahead and seek out the expected outcomes in the future. We then introduce the \roundinlinebox[green]{rewards} for assessing each state during reasoning in Section~\ref{sec:reward}. Guided by the world model and rewards, the planning with \roundinlinebox[green]{Monte Carlo Tree Search} (MCTS) efficiently explores the vast reasoning space and finds optimal reasoning traces (Section~\ref{sec:mcts}). Finally, when multiple promising reasoning traces are acquired during planning, we further introduce an aggregation method in Section~\ref{sec:aggr} that yields an ensembled result and further boosts the reasoning performance.

% which simulates each step of the reasoning trace, as described in Section \ref{sec:formulation}.
% This world model not only provides the outcome of the reasoning step but also evaluates its effectiveness, as described in Section \ref{sec:formulation}.
% We then discuss how to evaluate the effectiveness of reasoning steps to provide rewards in Section \ref{sec:reward}.
% Subsequently, we present our planning algorithm adapted from the Monte Carlo Tree Search (MCTS) algorithm to perform {planning} of the reasoning trace using the world model, as outlined in Section \ref{sec:mcts}.
% Finally, when multiple promising reasoning traces are explored during the planning process, we further introduce RAP-aggregate in Section \ref{sec:aggr} to select the best answer by aggregating these reasoning traces.

% drawing inspiration from the human ability to validate results by considering multiple reasoning traces~\cite{wang2022self}, we introduce RAP-aggregate in Section \ref{sec:aggr}. RAP-aggregate enables the selection of the best answer by aggregating the results obtained from multiple reasoning traces explored during the planning process.

% In this paper, we propose a new framework of LLM reasoning, namely \textbf{Reasoning via Planning (RAP)}, which leverage the LLM as both a world model and an agent.
% In Section \ref{sec:formulation}, we describe how to employ a LLM as the world model to explicitly predicts the next state and reward at each reasoning steps, transforming a reasoning task into a problem of planning.
% In Section \ref{sec:mcts}, we demonstrate how an agent LLM model, with the powerful feedback from the world model, can strategically plan a reasoning trace using the Monte Carlo Tree Search algorithm \cite{browne2012survey}.
% We finally introduce RAP-aggregate in Section \ref{sec:aggr}, which can aggregate a best answer across multiple reasoning traces we find during the planning.

%\vspace{-5pt}
\subsection{Language Model as World Model}
\label{sec:formulation}
%\vspace{-5pt}

%https://courses.cs.washington.edu/courses/cse599i/18wi/resources/lecture19/lecture19.pdf

% \begin{figure}[t]
%     \centering
%     \includegraphics[width=\textwidth]{sections/fig_formulation.pdf}
%     \caption{Formulation of RAP in an example from PrOntoQA. The left is the formulation of LLM as World Model, and the right is a search tree in this example.}
%     \label{fig:formulation}
% \end{figure}

In general, a world model predicts the next \emph{state} of the reasoning after \roundinlinebox[green]{applying an \emph{action} to the current state}~\cite{ha2018world, matsuo2022deep}. \ours enables us to instantiate the general concepts of state and action in different ways depending on the specific reasoning problems at hand (Figure \ref{fig:tree_examples}). For example, in \blocksworld, it is natural to define a state as the configuration of blocks (described in natural language), and an action to be a behavior of moving a block (e.g., ``pickup the orange block''). In a math reasoning problem, we use the state to represent the values of intermediate variables, and set an action to be a subquestion that drives the reasoning to derive new values. In logical reasoning, a state is a fact we are focusing on, and an action is to choose a rule for the next deduction.

With the definition of state and action, the reasoning process can thus be described as a \roundinlinebox[green]{Markov decision process (MDP):} given the current state $s_{t, t=0, 1, \dots, T}$, e.g., the initial state $s_0$, the LLM (as a reasoning agent) generates an action space by sampling from its generative distribution $a_t \sim p(a | s_t, c)$, where $c$ is a proper prompt (e.g., in-context demonstrations). Once an action is chosen, the world model then predicts the next state $s_{t+1}$ of the reasoning. Specifically, we repurpose the \emph{same} LLM to obtain a state transition distribution $p(s_{t+1} | s_t, a_t, c')$, where $c'$ is another prompt to guide the LLM to generate a state. For instance, in \blocksworld, the LLM (as the world model) generates text $s_{t+1}$ to describe the new configuration of blocks, given previous state $s_{t}$ and the action $a_t$.

Continuing the process results in a reasoning trace, which consists of a sequence of interleaved states and actions $(s_0, a_0, s_1, \dots, a_{T-1}, s_T)$. This differs from the previous reasoning methods, such as Chain-of-Thought~\cite{wei2022chain}, where the intermediate reasoning steps consist of only a sequence of actions, e.g., (\texttt{$a_0=$ ``pickup red block'', $a_1=$ ``stack on yellow block'',} \dots) (see comparisons in Figure~\ref{fig:1}). Augmenting the reasoning with the (predicted) world states helps the LLM with a more grounded and coherent inference. Note that the full reasoning trace is simulated by the LLM itself (as a reasoning agent with an \emph{internal} world model) without interacting with the \emph{external} real environment. \highlight[green]{This resembles humans contemplating a possible plan in their minds. The capability of simulating future states, by introducing the world model, allows us to incorporate principled planning algorithms to efficiently explore the vast reasoning space} as described in Section~\ref{sec:mcts}.

% To introduce the world model for simulating reasoning traces, we begin by defining the concepts of \textit{states} and \textit{actions} within a reasoning task.
% % We illustrate this with an example taken from the PrOntoQA dataset \cite{saparov2022language}, as shown in Figure \ref{fig:formulation}.
% We illustrate this with an example derived from the \blocksworld benchmark \cite{valmeekam2022large}.
% In our approach, we represent the reasoning process as a Markov decision process (MDP) with a state space $\mathcal S$ and an action space $\mathcal A$, as depicted in Figure \gy{ref figure}. %\ref{fig:formulation} (left).
% Each state $s \in \mathcal S$ corresponds to an intermediate status in the reasoning, specifically a language description of the current configurations of blocks in the \blocksworld context. %``\texttt{Fae is a cat}''.
% % At time $t$, an agent proposes an action $a_t \in \mathcal A$, which represents a rule to be employed in the subsequent reasoning step, such as ``\texttt{Every cat is a feline}'', based on the current state $s_t$ and its policy $p_\phi(a \mid s_t)$.
% At time $t$, an agent proposes an action $a_t \in \mathcal A$, which represents the subsequent reasoning step, such as a movement of a single block, based on the current state $s_t$ and its policy $p(a \mid s_t)$.
% The world model then conducts a one-hop reasoning step using $s_t$ and $a_t$, transitioning the state into the subsequent state $s_{t+1}$, which in this case denotes the new configuration of blocks after performing the movement $a_t$ on the current configuration $s_t$.
% The specific definition of states and actions can vary depending on the nature of the reasoning task.
% For example, in GSM8k \cite{cobbe2021training}, a dataset of numerical reasoning, the action is defined as proposing an incremental sub-question, while the state comprises the history of previous sub-questions and their corresponding answers.
% In terms of state transition, the world model attempts to answer the sub-question based on the problem context and the current state, subsequently incorporating the answer into the history to form the next state.

% In order to enable the world model to predict state transitions, which are subsequently utilized by agents for planning, we leverage the LLM as the state transition probability function $p_\theta$.
% In the \blocksworld example, we prompt the LLM the domain rule, the state $s_t$, the action $a_t$, and few-shot examples, to predict the next state $s_{t+1}$.

% In the PrOntoQA example, we prompt the LLM the state $s_t$ and action $a_t$ with few-shot examples to get the next state $s_{t+1}$.
% We also prompt it ``\texttt{Is this reasoning step correct?}'', and use the probability of ``\texttt{Yes}'' as the reward, evaluating the correctness of this reasoning step.

%However, the LLM often makes mistakes in retrieving accurate information or performing correct mathematical calculations.
%To ensure the reliability of our world model, we sample the LLM multiple times and select the next state with the highest confidence (The frequency of an answer). \gy{OK to directly say confidence?}
%We also use this confidence value as a reward, discouraging excessively difficulty or skipping sub-questions.

% In plan generation tasks, the world model is used to simulate actions by predicting the state transitions and the rewards without requiring direct access to the environment.
% In numerical and logical reasoning tasks, the LLM is utilized to answer sub-questions and perform atomic logical reasoning steps, allowing us to learn the outcome of reasoning steps and providing evaluations how good they are.
% Through the integration of the world model, a reasoning task is transformed into a strategic planning problem over the MDP, guided by the rewards.

% To address this, we employ a LLM as the world model, leveraging it to simulate the state transition function and the reward function.
% In abstract reasoning tasks \gy{unsure how to describe this type}, the definition of states and actions are more flexible.
% A partial solution $s \in \mathcal S$ can correspond to an intermediate conclusion or the previous answers of decomposed sub-questions. A reasoning step $a \in \mathcal A$ can refer to a given fact or a new sub-question that leads to the next state.
% In these cases, we also utilize a LLM as the world model to simulate the state transition, enabling us to derive the next intermediate conclusion or answer the sub-question. We also evaluate the quality of each reasoning step using the LLM as the reward function.
% We illustrate the detailed MDP formulations of Blocksworld, ProntoQA, and GSM8k in \gy{table/figure}. \gy{we want to include action proposing in world model, need modify}

% We assume the MDP has finite horizon, which means any transition sequence of nonzero probability (i.e. $ \prod_{i \geq 0} \mathcal{P}\left(x_i, a_{i, i+1}, x_{i+1}\right)>0$) must have finite length.
% Since their concrete interpretations of the MDP depend on specific problems, we leave them to later sections \hsb{ref}. Here we take a case from blockworld as a running example: \hsb{expand}

% In our method, the transition probability function and the reward function are parameterized by the world model, noted as $\mathcal{P_\phi}$ and $\mathcal{R_\phi}$. \hsb{example}

% A planning problem is formulated as below:

%\vspace{-5pt}
\subsection{Reward Design} \label{sec:reward}
%\vspace{-5pt}

During reasoning, we want to assess the feasibility and desirability of each reasoning step, and guide the reasoning based on the assessment (Section~\ref{sec:mcts}).
The assessment of each reasoning step (i.e., applying an action $a_t$ to the state $s_{t}$) is performed by a \emph{reward} function $r_t = r(s_t, a_t) \in \mathbb R$. Similar to the state and action, the reward function can be specified in different ways to accommodate any knowledge or preferences about the reasoning problem of interest. Here we introduce several common rewards applicable to different tasks and shown to be effective in our experiments.

\noindent \textbf{Likelihood of the action.}
When an action is generated by the LLM conditioning on the in-context demonstration and the current state, the probability of the specific action reflects the LLM's preference. We thus can incorporate the log probability of the action as a reward. This reward reflects the ``instinct'' of LLMs as an agent, and can be also used as a prior for which action to explore.

\noindent \textbf{Confidence of the state.}
State prediction is nontrivial in some problems, e.g., in math reasoning (Figure~\ref{fig:tree_examples}, middle), given an action (i.e., a subquestion), the world model predicts the next state by answering the subquestion. We incorporate the confidence of the state (i.e., answers in this case) as a reward. Specifically, we draw multiple sample answers from the world model, and use the proportion of the most frequent answer as the confidence. Higher confidence indicates that the state prediction is more consistent with the world knowledge of LLMs \cite{hao2023bertnet}, which typically leads to a more reliable reasoning step.

% In certain tasks, individual reasoning step can still be challenging.
% For example, in the GSM8k dataset \cite{cobbe2021training}, a reasoning step involves proposing (action) and answering a sub-question (state transition) based on previous information.
% We then use the confidence of answers as a reward. Specifically, we sample multiple responses from the world model and group them by the answer. The maximal frequency of an answer is defined as confidence, and high confidence indicates reliable reasoning steps.

% \textbf{Correctness estimation by LLM. \gy{Since we use this way to calculate correctness in PrOntoQA and helpfulness in GSM8k, shall we replace with `Self-evaluation by LLM'} \hzt{sounds good. and please update the paragraph}}
\noindent \textbf{Self-evaluation by the LLM.}
It's sometimes easier to recognize the errors in reasoning than avoid generating them in advance. Thus, it's beneficial to allow the LLM to criticize itself with the question ``\texttt{Is this reasoning step correct?}'', and use the next-word probability of the token ``\texttt{Yes}'' as a reward. The reward evaluates LLM's own estimation of the correctness of reasoning. Note that the specific problems for self-evaluation can be different depending on the tasks.

% Likewise, we can prompt the LLM with the question ``\texttt{Is this reasoning step correct?}'' and utilize the probability of the token \texttt{Yes} as a reward, reflecting the LLM's evaluation of the reasoning step's reliability,

\noindent \textbf{Task-specific heuristics.}
RAP also allows us to flexibly plug in other task-specific heuristics into the reward function. For example, in plan generation for \blocksworld, we compare the predicted current state of blocks with the goal to calculate a reward (Section~\ref{sec:plan}). The reward encourages the plan of movements to actively pace towards the target.

% In plan generation tasks, we can also borrow the rewards from the corresponding environment. For example, in \blocksworld benchmark \cite{valmeekam2022large}, we can assign reward by the number of sub-goals that are fulfilled when performing the movement in action $a_t$.

% As a by-product, the new state is predicted as the ensemble of multiple samples, which may better align with the LLM's belief about the world \cite{jung2022maieutic, hao2022bertnet}.

% Besides the world model, we define a reward function $r_t = r(s_t, a_t) \in \mathbb R$ for each task to assess the quality of this reasoning step.
% The reward function can incorporate various evaluation approaches, some of which are outlined below.

% \vspace{-5pt}
\subsection{Planning with Monte Carlo Tree Search} \label{sec:mcts}
% \vspace{-5pt}

\begin{figure*}[t]
    \centering
    \includegraphics[width=\textwidth]{sections/mcts_phases.pdf}
    \vspace{-25pt}
    \caption{An illustration of the four phases in an iteration in MCTS planning (Section~\ref{sec:mcts}).}
    \label{fig:mcts_phases}
    \vspace{-12pt}
\end{figure*}

Once equipped with the world model (Section~\ref{sec:formulation}) and rewards (Section~\ref{sec:reward}), LLMs can reason with any planning algorithms. We adopt Monte Carlo Tree Search (MCTS)~\cite{kocsis2006bandit, coulom2007efficient}, a powerful planning algorithm that strategically explores the space of reasoning trees and strikes a proper balance between exploration and exploitation to find high-reward reasoning traces efficiently.

Specifically, MCTS builds a reasoning tree iteratively, where each node represents a state, and each edge represents an action and the transition from the current state to the next state after applying the action (Figure~\ref{fig:1}).
To guide the LLM agent to expand and explore the most promising nodes of the tree, the algorithm maintains a state-action value function $Q : \mathcal S \times \mathcal A \mapsto \mathbb R$, where $Q(s, a)$ estimates the \emph{expected future reward} of taking action $a$ in state $s$. Figure \ref{fig:mcts_phases} illustrates four operations in each iteration to expand the tree and update $Q$ values. The process continues until a specified computational budget (e.g., the number of iterations) is reached, and the resulting traces are acquired from the tree. More details and the pseudo-code of the planning algorithm are given in Appendix~\ref{sec:mcts_app} and Algorithm \ref{alg:mcts}.

% That is, we assess the potential of a node (a reasoning step) by looking ahead and anticipating the reward in future trajectories starting from this node. This fundamentally differs from the current reasoning methods that generate a reasoning trace autoregressively without considering the future.

% More specifically, as illustrated in Figure \ref{fig:mcts_phases}, the MCTS planning performs four operations in each iteration to expand the tree and update $Q$ values, i.e., \textit{selection}, \textit{expansion}, \textit{simulation}, and \textit{back-propagation}. The process continues until a specified computational budget (e.g., the number of iterations) is reached, and the resulting reasoning traces are acquired from the tree, as we articulated later. The psuedo-code of our MCTS planning is given in Algorithm \ref{alg:mcts} in the Appendix.
% The iterations continue until a target state is reached, resulting in a complete reasoning trace for the target problem.

% We employ Monte Carlo Tree Search (MCTS), a decision-making algorithm that explores the MDPs using a decision tree, by which we can balance exploration and exploitation when planning the reasoning trace. In the decision tree representation, each node represents a state, while each edge represents an action and the corresponding state transition, as illustrated in Figure \ref{fig:formulation} (right).

% The algorithm maintains an optimal state-action value function $Q : \mathcal S \times \mathcal A \mapsto \mathbb R$, where $Q(s, a)$ estimates the maximum expected reward starting from state $s$ with action $a$.
% MCTS operates iteratively by sampling a complete trajectory (i.e., a reasoning trace) from the root of the tree (initial state $s_0$) towards a terminating node, where the agent finishes the reasoning.
% This iterative process expands the decision tree incrementally with each iteration.
% Additionally, the $Q$ values of the state-action pairs along the trajectory are updated in each iteration.
% Every iteration is composed four phases, \textit{selection}, \textit{expansion}, \textit{simulation}, and \textit{back-propagation}, demonstrated in Algorithm \ref{alg:mcts} with details below.

% In reasoning tasks, the state space and action space of half-way solutions can be very large and even infinite due to the flexibility to propose reasoning steps in natural language, which means that the state space cannot be fully searched.
% In order to do planning on the MDP within a computational budget, we introduce the Monte Carlo Tree Search algorithm \cite{browne2012survey} to strategically and efficiently search for a good plan.
% The algorithm builds a rooted tree of the search history, where each node represents a state, while each edge represents an action.
% It also maintains an optimal state-action value function $Q: \mathcal S \times \mathcal A \mapsto \mathbb R$, which will be used in roll-out policy.
% The algorithm will do several iterations, called roll-outs, each of which consists of the following four phases.
% We present the roll-out procedure in Algorithm \ref{alg:mcts} with psuedo-code.

\noindent \textbf{Selection.}
The first phase selects a portion of the existing tree that is most promising for further expansion in the next phase. Starting from the root node (i.e., initial state $s_0$), at each level of the tree, the algorithm selects a child node as the next node. The phase finishes when a leaf node of the current tree is reached. Figure~\ref{fig:mcts_phases}(a) highlights the selected path in red. To balance between exploration (of less-visited nodes) and exploitation (of high-value nodes), we use the well-known \emph{Upper Confidence bounds applied to Trees (UCT)} \cite{kocsis2006bandit} to select each child node. Specifically, at node $s$, we select the action in the tree by considering both the $Q$ value (for exploitation) and uncertainty (for exploration):

\vspace{-8pt}
{
\small
\begin{align}
    a^\ast = \arg\max_{a \in A(s)} \left[ Q(s, a) + w \sqrt{\frac{\ln N(s)}{N(c(s, a))}} \right],
\end{align}
}
\noindent where $N(s)$ is the number of times node $s$ has been visited in previous iterations, and $c(s, a)$ is the child node of applying $a$ in state $s$. The less a child node was visited before (i.e., the more uncertain about this child node), the higher the second term in the equation. The weight $w$ controls the balance between exploration and exploitation.
% \hzt{Anything to say about how we chose $w$?}

% Each iteration starts the initial state.
% In each step, the algorithm will select a child node based on a \textit{tree policy}, until reaching an unvisited node.
% To balance between exploration and exploitation, we use the maximum Upper Confidence Trees (UCT) policy, which select the child node based on maximizing upper confidence.
% The upper confidence is determined by the summation of the $Q$ value, which encourages exploitation, and an uncertainty term, which encourages exploration.
% \begin{align}
%     a^\ast = \arg\max_{a \in A(s)} \left[ Q(s, a) + w \sqrt{\frac{\ln N(s)}{N(c(s, a))}} \right],
% \end{align}
% where $N(s)$ is the number of times node $s$ has been visited, and $c(s, a)$ is the children of following $a$ at node $s$.
% The hyper-parameter $w$ serves as a balancing factor between exploration and exploitation.

\noindent \textbf{Expansion.}
This phase expands the tree by adding new child nodes to the leaf node selected above. Given the state of the leaf node, we use the LLM (as agent) to sample $d$ possible actions (e.g., subquestions in math reasoning), and then use the LLM (as world model) to predict the respective next state, resulting in $d$ child nodes. Note that if the leaf node selected above is a terminal node (the end of a reasoning chain) already, we will skip expansion and jump to back-propagation.
% Note that if the leaf node selected above is a terminal (target) state already, we will skip expansion/simulation and jump to back-propagation.

% When reaching a node that have never been visited, we add to the tree the children of the node.
% Since the space of language representations can be very large, we sample $d$ possible actions using the LLM as a substitute action space $A(s)$, and add the $d$ corresponding children to the tree.

% Since the space of action, which is a language representation of the next reasoning step, can be very large, we sample $d$ possible actions using the world model LLM as the children, where $d$ is a pre-defined hyper-parameter.
% To simplify the problem, we sample a next state using the transition probability function, and reuse the sample in the future.
% The difficulty of predicting state transitions vary among tasks.
% For easier transitions, we to generate the next state with greedy decoding, while for harder transitions, we utilize self-consistency, which means we use the next state with highest confidence across several generations.
% We will also have a quick evaluation of each action, namely a prior $r'$, which can be but not necessarily be the reward defined in the world model.
% We use $r'$ to temporarily serve as $Q(s, a)$ before a roll-out actually follows $a$ at the node representing the state $s$.

\noindent \textbf{Simulation.}
To estimate the expected future rewards ($Q$ values), this phase simulates the future situations of the current node using the world model.
% Specifically,
%from the above $d$ new nodes, we pick the node of largest local reward (Section~\ref{sec:reward}) and perform simulation starting from it. In particular, similar to above,
Starting from the current node as above, at each node $s$, we create an action following a \emph{roll-out policy} and use the world model to predict the next state. The roll-out process continues until a terminal state is reached. There could be many ways to define the roll-out policy (e.g., by adding different randomness). In our experiments, for simplicity and reduced noises, we follow a similar process as in the expansion above, i.e., generating $d$ candidate actions and picking one of the largest local reward $a'= \max_{a'} r(s, a)$. In practice, for efficiency, we discard the computationally costly components in $r$ (e.g., the reward from the confidence of state requires sampling the answer multiple times), and use the resulting lightweight reward function for selecting actions during simulation.

% In traditional MCTS algorithms, after expanding the children of one node, it will follow a random simulation till a terminating state, i.e., a leaf node.
% To encourage efficient exploration, we instead follow the children with the highest prior $r'$, and repeat this procedure after expanding the new visited node till the end.

% After expansion, we perform a complete simulation starting from the current node until the reasoning trace is finished following a \textit{roll-out policy}. To guide our exploration towards better reasoning traces, we use a fast reward function $r'(s, a)$, discarding time-consuming parts of the standard reward function $r(s, a)$, to get an estimation of the quality of action $a$. Our roll-out policy selects the action with maximum estimated reward $a_t = \max_{a_t} r'(s, a_t)$.

\noindent \textbf{Back-propagation.}
Once we reach a terminal state in the above phases, we obtain a reasoning path from the root node to the terminal node. We now back-propagate the rewards on the path to update the $Q$ value of each state-action pair along the path. Specifically, we update $Q(s, a)$ by aggregating the rewards in all future steps of node $s$.
% We may adopt the aggregation method depending on the nature of tasks and reward design, as discussed in Section~\ref{sec:experiments}.
% After an iteration reaches a terminal state in the above phases.
% We back propagated the rewards on the path, updating the $Q$ function with each state-action pair in the trajectory with rewards in all future steps.
% Since the $Q$ function depends on the reward design, we will introduce in detail in Section \ref{sec:experiments}.

Once a predetermined number of MCTS iterations is reached, we terminate the algorithm and select the final reasoning trace from the constructed tree for evaluation. There are various ways for the selection.
% Following a predetermined number of iterations, there are several strategies for selecting the final reasoning path.
One is to start from the root node and iteratively choose the action with the highest $Q$ value until reaching a terminal.
Also, one can directly select the path from the iterations that yielded the highest reward, or opt to choose the leaf node (and the respective root-to-leaf path) that has been visited the most.
In practice, we observed that the second strategy often yields the best results.

% When solving a reasoning problem in the task, we deploy an agent to search over the corresponding MDP.
% In order to keep a track of the searching history, we build a reasoning tree representing the progress of searching.
% Each node on the tree represents a state $S \in \mathcal S$ in the MDP, where the root represents the initial state $S_0$.
% Every time the agent reaches a new node $S_t$ on the tree,

%\vspace{-5pt}
\subsection{RAP-Aggregation} \label{sec:aggr}
%\vspace{-5pt}

% : Aggregating Multiple Reasoning Outputs

For problems, such as math reasoning (Section~\ref{sec:math}) where only the final answer is required, \ours could produce multiple traces and answers from different MCTS iterations, which will be aggregated to produce the final answer. We refer to such a mechanism as \ours-Aggregation. Note that problems like plan generation or logical inference require a complete reasoning trace as output; thus, \ours-Aggregation will not be applied.

% More importantly, there is a concern that some incorrect reasoning steps may appear in the early stage of multiple iterations, thus polluting the aggregation. As a result, we further devise a new weighting strategy for aggregating candidate answers. Specifically, for each candidate answer, we accumulate the reward of each reasoning step in the answer's reasoning traces. We choose the answer with the highest accumulative reward as the final aggregated answer.

% Some of the reasoning problems, such as plan generation (Section~\ref{sec:plan}) and logical inference (Section~\ref{sec:logical}), require a complete reasoning trace as output. In other problems, such as math reasoning (Section~\ref{sec:math}), only the final answer is required. In such cases, we could collect multiple reasoning traces during RAP and aggregate the respective answers to produce the final answer, similar to the self-consistency method \cite{wang2022self} in Chain-of-Thought.
% Specifically, we collect reasoning traces and answers from different MCTS iterations. To address the concern that some incorrect early reasoning steps may appear in multiple iterations and thus pollute the aggregation, we devise a new  weighting strategy for aggregating candidate answers. Specifically, for each candidate answer, we accumulate the reward of each reasoning step in the answer's reasoning traces. We choose the answer with the highest accumulative reward as the final aggregated answer.

% However, in certain tasks such as many numerical reasoning tasks, only the final answer needs to be provided.
% When tackling numerical reasoning problems, humans often consider multiple reasoning paths to verify the answer further.
% Inspired by this, we introduce \textbf{RAP-aggregation}, where we aggregate the answer across iterations to enhance the output of MCTS.
% To address the concern that an early incorrect reasoning step may contribute multiple times to the final result when it appears in multiple iterations, we perform the aggregation in a basis of edges.
% Specifically, for a given potential answer, we accumulate the weights of all edges involved in at least one iteration leading to that particular answer.
% The weight of an edge is the reward assigned to its represented action.
% Ultimately, the answer with the highest total weight is selected as the aggregated output.
% To avoid an early incorrect reasoning step contributing multiple times to the result when multiple iterations include that step, we conduct the vote by edges.
% More specifically, for one possible answer, all edges involving in at least one iteration with that final answer will be added to the weight for the answer, and the weight is the reward of the action represented by the edge.

% Formally, we choose the answer $o$ following
% \begin{align}
%     o^\ast = \arg\max_{o} \sum_{e = (s, a): \exists T: T\text{~outputs~}o \wedge e \in T} r(s, a),
% \end{align}
% where $e$ is an edge on the tree representing following action $a$ when visiting the node representing $s$, and $T$ is the trace in one roll-out.

### 2024-07-07 nlp-papers/reasoning-via-planning/sections/method.tex
Add LLaMa downliad links

@@
 
-With the definition of state and action, the reasoning process can thus be described as a \roundinlinebox[green]{Markov decision process (MDP):} given the current state $s_{t, t=0, 1, \dots, T}$, e.g., the initial state $s_0$, the LLM (as a reasoning agent) generates an action space by sampling from its generative distribution $a_t \sim p(a | s_t, c)$, where $c$ is a proper prompt (e.g., in-context demonstrations).
+With the definition of state and action, the reasoning process can thus be described as a \highlight[green]{Markov decision process (MDP): given the current state $s_{t, t=0, 1, \dots, T}$, e.g., the initial state $s_0$, the LLM (as a reasoning agent) generates an action space by sampling from its generative distribution $a_t \sim p(a | s_t, c)$, where $c$ is a proper prompt (e.g., in-context demonstrations).
 Once an action is chosen, the world model then predicts the next state $s_{t+1}$ of the reasoning.
 Specifically, we repurpose the \emph{same} LLM to obtain a state transition distribution $p(s_{t+1} | s_t, a_t, c')$, where $c'$ is another prompt to guide the LLM to generate a state.
-For instance, in \blocksworld, the LLM (as the world model) generates text $s_{t+1}$ to describe the new configuration of blocks, given previous state $s_{t}$ and the action $a_t$.
+For instance, in \blocksworld, the LLM (as the world model) generates text $s_{t+1}$ to describe the new configuration of blocks, given previous state $s_{t}$ and the action $a_t$}.
 
